Illuminating a Dark Lens : A Type Ia Supernova \\ Magnified by Galaxy Cluster Abell 2744

### 2015-04-07 TABLE/magnifications.tex
incorporated AZ comments

\begin{deluxetable*}{lccp{1.9in}p{2.2in}}

@@

\tablehead{ \colhead{Model} & \colhead{Best}\tablenotemark{a} & \colhead{Median\tablenotemark{b}} & \colhead{References} & \colhead{Description}}

@@

Sharon(v1)   & 2.53 & 2.57$^{+0.18}_{-0.16}$ & \citealt{Jullo:2007}  & LENSTOOL parametric, strong-lensing based model\\
Sharon(v2)   & 2.73 & 2.69$^{+0.14}_{-0.06}$ &   \citealt{Jullo:2007};\citealt{Johnson:2014} & LENSTOOL parametric, strong-lensing based model\\
CATS-SL      & 2.25 & 2.27$^{+0.05}_{-0.04}$ &   \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL parametric strong-lensing based model.\\
CATS-SL+WL   & \nodata & 2.62$^{+0.18}_{-0.18}$ & \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL parametric model with both strong and weak lensing constraints.\\
Jauzac		 & \nodata & 3.37$^{+0.14}_{-0.15}$ &   \citealt{Jauzac:2014,Richard:2014} & Updated version of the CATS-SL model, adds 33 new multiply-imaged galaxies.\\
GLAFIC       & 2.32 & 2.28$^{+0.07}_{-0.11}$ &   \citealt{Oguri:2010,Ishigaki:2015} & Parametric strong-lensing model using the {\tt GLAFIC} code.\tablenotemark{d} \\
Zitrin-NFW   & 2.07 & 2.27$^{+0.23}_{-0.22}$ &   \citealt{Zitrin:2009a} &  Parametric strong-lensing model using PIEMD\tablenotemark{e} profiles for galaxies and NFW\tablenotemark{f} profiles for dark matter halos.\\
Zitrin-LTM   & 2.64 & 2.96$^{+0.77}_{-0.38}$ &   \citealt{Zitrin:2013a} & Parametric strong-lensing model, adopts the Light-Traces-Mass assumption for both the luminous and dark matter.\\
Bradac(v1)   & 3.15 & 2.45$^{+0.19}_{-0.16}$ &   \citealt{Bradac:2005,Bradac:2009} & {\tt SWUnited} : Free-form, strong+weak-lensing based model. Errors from bootstrap resampling only weak-lensing constraints.\\
Bradac(v2)   & 2.21 & 2.23$^{+0.05}_{-0.03}$ &   \citealt{Wang:2015} & \change{Updated version of the {\tt SWUnited} model with new strong-lensing constraints from HFF imaging. Errors from bootstrap resampling only strong-lensing constraints.}\\
Williams     & 2.67 & 2.78$^{+2.68}_{-1.14}$ &   \citealt{Liesenborgs:2006,Liesenborgs:2007,Mohammed:2014} & {\tt GRALE}\tablenotemark{g} : Free-form strong-lensing model using a genetic algorithm.  \\
Merten       & 2.31 & 2.22$^{+0.67}_{-0.19}$ &   \citealt{Merten:2009,Merten:2011} &  {\tt SaWLENS},\tablenotemark{h} Grid-based free-form strong+weak lensing based model using adaptive mesh refinement.\\
Lam			 & \nodata & 2.77$^{+0.36}_{-0.36}$ &   \citealt{Sendra:2014,Lam:2014} & {\tt WSLAP+}\tablenotemark{h} : Free-form model strong-lensing using a grid-based method, supplemented by deflections fixed to cluster member galaxies.\\
Diego\tablenotemark{i}			 & \nodata & 2.10$^{+0.36}_{-0.36}$ &   \citealt{Sendra:2014,Lam:2014} & \change{Alternative implementation of the {\tt WSLAP+}\tablenotemark{h} model, using a different set of strong-lensing constraints and redshifts.}

@@

\tablenotetext{b}{Median magnification from 100-600 Monte Carlo realizations of the model. Uncertainties enclose 68\%\ of the realized values.}  

@@

\tablenotetext{i}{\change{No uncertainty estimates were available for the Diego implementation of the {\tt WSLAP+} model, so we adopt the uncertainties from the Lam model.}}

>>>>

\begin{deluxetable*}{lcccp{1.4in}p{2.1in}}

@@

\tablehead{ \colhead{Model} & \colhead{Best\tablenotemark{a}} & \colhead{Median\tablenotemark{b}} & \colhead{68\% Conf. Range\tablenotemark{b}} & \colhead{References} & \colhead{Description}}

@@

Sharon(v1)   & 2.53     & 2.57  &   2.41$-$2.75 & \citealt{Jullo:2007}  & LENSTOOL parametric, strong-lensing based model\\
Sharon(v2)   & 2.73     & 2.69  &   2.63$-$2.83 &   \citealt{Jullo:2007};\citealt{Johnson:2014} & LENSTOOL parametric, strong-lensing based model.\\
CATS-SL      & 2.25     & 2.27  &   2.23$-$2.32 &   \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL parametric strong-lensing based model.\\
CATS-SL+WL   & \nodata  & 2.62  &   2.44$-$2.80 & \citealt{Jullo:2009,Jauzac:2012} &  LENSTOOL parametric model with both strong and weak lensing constraints.\\
Jauzac		 & \nodata  & 3.37  &   3.22$-$3.51 &   \citealt{Jauzac:2014,Richard:2014} & Updated version of the CATS-SL model, adds 33 new multiply-imaged galaxies.\\
GLAFIC       & 2.32     & 2.28  &   2.17$-$2.35 &   \citealt{Oguri:2010,Ishigaki:2015} & Parametric strong-lensing model using the {\tt GLAFIC} code.\tablenotemark{d} \\
Zitrin-NFW   & 2.07     & 2.27  &   2.05$-$2.50 &   \citealt{Zitrin:2013a} &  Parametric strong-lensing model using PIEMD\tablenotemark{e} profiles for galaxies and NFW\tablenotemark{f} profiles for dark matter halos.\\
Zitrin-LTM   & 2.64     & 2.96  &   2.58$-$3.73 &   \citealt{Zitrin:2009a} & Parametric strong-lensing model, adopts the Light-Traces-Mass assumption for both the luminous and dark matter.\\
Bradac(v1)   & 3.15     & 2.45  &   2.29$-$2.64 &   \citealt{Bradac:2005,Bradac:2009} & {\tt SWUnited} : Free-form, strong+weak-lensing based model. Errors from bootstrap resampling only weak-lensing constraints.\\
Bradac(v2)   & 2.21     & 2.23  &   2.20$-$2.28 &   \citealt{Wang:2015} & \change{Updated version of the {\tt SWUnited} model with new strong-lensing constraints from HFF imaging. Errors from bootstrap resampling only strong-lensing constraints.}\\
Williams     & 2.67     & 2.78  &   1.64$-$5.46 &   \citealt{Liesenborgs:2006,Liesenborgs:2007,Mohammed:2014} & {\tt GRALE}\tablenotemark{g} : Free-form strong-lensing model using a genetic algorithm.  \\
Merten       & 2.31     & 2.22  &   2.03$-$2.89 &   \citealt{Merten:2009,Merten:2011} &  {\tt SaWLENS},\tablenotemark{h} Grid-based free-form strong+weak lensing based model using adaptive mesh refinement.\\
Lam			 & \nodata  & 2.77  &   2.41$-$3.13 &   \citealt{Sendra:2014,Lam:2014} & {\tt WSLAP+}\tablenotemark{h} : Free-form model strong-lensing using a grid-based method, supplemented by deflections fixed to cluster member galaxies.\\
Diego\tablenotemark{i}        & \nodata  & 2.10  &   1.74$-$2.46 &   \citealt{Sendra:2014,Lam:2014} & \change{Alternative implementation of the {\tt WSLAP+}\tablenotemark{h} model, using a different set of strong-lensing constraints and redshifts.}

@@

\tablenotetext{b}{Median magnification from 100-600 Monte Carlo realizations of the model. Confidence ranges enclose 68\%\ of the realized values.}  

@@

\tablenotetext{i}{\change{No uncertainty estimates were available for the Diego implementation of the {\tt WSLAP+} model, so we adopt the uncertainties from the closely related Lam model.}}

### 2015-04-07 snTomas.tex
incorporated AZ comments

All of the imaging data were processed using the {\tt sndrizpipe}
pipeline,\footnote{\url{https://github.com/srodney/sndrizpipe} v1.2
DOI:10.5281/zenodo.10731} a custom data reduction package in Python
that is employs the {\tt DrizzlePac} tools from the Space Telescope
Science Institute (STScI) \citep{Fruchter:2010}.  Photometry was
collected using the {\tt PyPhot} software
package,\footnote{\url{https://github.com/djones1040/PyPhot}} a
pure-Python implementation of the photometry algorithms from the IDL
AstroLib package \citep{Landsman:1993}, which in turn are based on the
DAOPHOT program \citep{Stetson:1987}.  For the IR bands we used point
spread function (PSF) fitting on the difference images, and in the ACS
optical bands we collected photometry with a
0\farcs3 aperture. Table~\ref{tab:Photometry} presents the list of
observations, along with measured photometry from all available
imaging data.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_lightcurve_classification}
\caption{ 
Maximum likelihood model for each SN sub-class, derived from Bayesian
model selection using the photometric data alone. Grey points show the
observed SN \tomas\ photometry with error bars, though these are typically
smaller than the size of the marker.  The Type Ia model
(orange solid line) is drawn from the SALT2 template at $z=1.35$. The
best match from all Type Ib/Ic models is based on the Type Ic SN
SDSS-14475 at $z=0.695$ (green dashed line). For the Type II class, the
best match is from the Type II-L SN 2007pg at $z=1.8$ (blue dash-dot line). The Type
Ia model is by far the best match, and the only one that is consistent
with both the photo-$z$ of the probable host galaxy and the
spectroscopic redshift from the SN spectrum.
\label{fig:photoclass} }
\end{center}
\end{figure}

As a check on the spectral classification of SN \tomas\
(Section~\ref{sec:SNID}), we independently classified the SN using a
Bayesian photometric classifier.  We use the
{\tt sncosmo} software
package\footnote{\url{http://sncosmo.github.io/}} to simulate SN light
curves from $z=0.3$ to 2.3 and evaluate the classification
probability using traditional Bayesian model selection \citep[as
in][]{Jones:2013,Rodney:2014,Graur:2014,Rodney:2015a}.  In this
analysis we represent normal Type Ia SNe with the SALT2
model \citep{Guy:2010}, and \CCSNe\ with 42 discrete templates (26
Type II and 16 Type Ib/c) drawn from the template library of the
SuperNova Analysis software
package \citep[SNANA,][]{Kessler:2009a}.\footnote{Throughout this work
we use SNANA v10\_35g.}
Likelihoods are defined by comparing the observed fluxes to model
predictions in all passbands where the model is defined.  In practice,
this means we exclude the F435W and F606W bands, which are too blue
for our models at $z>0.85$.

@@

The models shown in Figure~\ref{fig:lensingtest} are separated along
the $y$ direction into two broad classes, with 
eight \change{so-called} {\it parametric}\footnote{\change{Although this
nomenclature is becoming standard in the literature, it is somewhat
misleading, as the pixels or grid cells in free-form models are
effectively parameters as well. Perhaps ``simply-parameterized'' would
be more accurate, though we adopt the more common usage here.}}
models on top and six {\it free-form} models on the bottom.
Broadly speaking, the parametric models use parameterized density
distributions to describe the arrangement of mass within the cluster,
typically associating massive dark matter halos with the positions of
bright cluster member galaxies. Therefore, parametric models rely (to
varying degrees) on the assumption that the observed light from
cluster galaxies is a good tracer of the cluster's dark matter.
Free-form models divide the cluster into a grid, generally using a
\change{multi-scale grid} to get better sampling in denser regions. Each grid cell
is assigned a mass or a potential, and then the mass values are
iteratively refined to match the observed lensing
constraints. \change{In some cases an adaptive grid is used so
that the grid spacing itself can also be modified as the model is
iterated \citep[e.g.][]{Liesenborgs:2006,Merten:2009}.}

\change{
In general, free-form methods tend to explore the model parameter
space more fully than parametric methods.  This is reflected in the
size of the error-bars in Figure~\ref{fig:lensingtest}.  For example,
the Williams model (using GRALE), which has the largest, and possibly the
most realistic uncertainties is able to account for mass sheet
degeneracy by introducing a free parameter in the reconstruction that
represents an arbitrarily positioned mass sheet, with normalization to
be chosen by the genetic algorithm.  More complex degeneracies are
also accounted for in this model through the use of basis
functions \citep{Liesenborgs:2006,Liesenborgs:2007,Mohammed:2014} }

@@

This shift to higher magnifications is not, however, unique to the
SN \tomas\ position.  \citet{Jauzac:2014c} noted that their updated
mass model of Abell 2744 results in a systematic increase in the
magnification values across the cluster field. Examining a sample of
$\sim$30 multiple images, \citeauthor{Jauzac:2014c} found that the
magnifications increased by a factor of $\sim1.5$ relative to pre-HFF
models.  Although this single sight-line can only provide a very
limited test of the model, the \changeG{$4.3\sigma$} discrepancy is at
least a strong suggestion that improving the number and quality of
strong-lensing constraints can still lead to magnification maps that
are susceptible to systematic biases.

@@

The other three models in our comparison set also use the strong
lensing constraints from the core region, but additionally utilize
weak lensing measurements to provide additional constraints on the
mass distribution farther from the cluster core.  The signal from weak
lensing relies on a large sample of background galaxies, so this
constraint operates principally at separations more than 1\arcmin\ from
the cluster core.  At a projected separation of $\sim40\arcsec$,
SN \tomas\ falls in between the strong- and weak-lensing regimes, and
one would expect that models incorporating both of those constraints
would be less susceptible to the systematic bias of a sharply
steepening mass profile.  The lowest magnification estimate in our
sample (from the Merten model) is among this sub-sample.  However, we
find that collectively these three strong+weak models (shown as
diamonds in Figure~\ref{fig:lensingtest}) exhibit the same propensity
to overestimate the magnification along this line of sight.  Thus, a
sudden change in the mass profile outside the strong lensing region is
not a likely explanation for this small systematic bias.

@@

This single object behind a single cluster is not in and of itself a
cause for alarm.  Previous analyses of lensed \SNIa\ found no
significant discrepancy beween the observed \SNIa\ magnifications and
the predictions from lens models
(\citetalias{Patel:2014}; \citealt{Nordin:2014}), albeit with a much
smaller set of lens models being tested.  The observed systematic bias
for \tomas\ is small, and many of the lens models being evaluated are
preliminary models that have not been updated to include all of the
HFF data.  Future revisions of the lens models for Abell 2744 should
either incorporate the observed magnification of \tomas\ as a new
model constraint, or can revisit this test to evaluate whether the
bias persists.

>>>>

All of the imaging data were processed using the {\tt sndrizpipe}
pipeline,\footnote{\url{https://github.com/srodney/sndrizpipe} v1.2
DOI:10.5281/zenodo.10731} a custom data reduction package in Python
that employs the {\tt DrizzlePac} tools from the Space Telescope
Science Institute (STScI) \citep{Fruchter:2010}.  Photometry was
collected using the {\tt PyPhot} software
package,\footnote{\url{https://github.com/djones1040/PyPhot}} a
pure-Python implementation of the photometry algorithms from the IDL
AstroLib package \citep{Landsman:1993}, which in turn are based on the
DAOPHOT program \citep{Stetson:1987}.  For the IR bands we used point
spread function (PSF) fitting on the difference images, and in the ACS
optical bands we collected photometry with a
0\farcs3 aperture. Table~\ref{tab:Photometry} presents the list of
observations, along with measured photometry from all available
imaging data.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{FIG/snTomas_lightcurve_classification}
\caption{ 
Maximum likelihood model for each SN sub-class, derived from Bayesian
model selection using the photometric data alone. Grey points show the
observed SN \tomas\ photometry with error bars, though these are
typically smaller than the size of the marker.  The Type Ia model
(orange solid line) is drawn from the SALT2 template at $z=1.35$. The
best match from all Type Ib/Ic models is based on the Type Ic SN
SDSS-14475 at $z=0.695$ (green dashed line). For the Type II class,
the best match is from the Type II-L SN 2007pg at $z=1.8$ (blue
dash-dot line). The Type Ia model is by far the best match, and the
only one that is consistent with both the photo-$z$ of the probable
host galaxy and the spectroscopic redshift from the SN spectrum.  The
date of the HST spectral observations is marked with a thin grey
vertical line.
\label{fig:photoclass} }
\end{center}
\end{figure}

As a check on the spectral classification of SN \tomas\
(Section~\ref{sec:SNID}), we independently classified the SN using a
Bayesian photometric classifier.  We use the {\tt sncosmo} software
package\footnote{\url{http://sncosmo.github.io/}} to simulate SN light
curves from $z=0.3$ to 2.3 and evaluate the classification probability
using traditional Bayesian model selection \citep[as
in][]{Jones:2013,Rodney:2014,Graur:2014,Rodney:2015a}.  In this
analysis we represent normal Type Ia SNe with the SALT2
model \citep{Guy:2010}, and \CCSNe\ with 42 discrete templates (26
Type II and 16 Type Ib/c) drawn from the template library of the
SuperNova Analysis software
package \citep[SNANA,][]{Kessler:2009a}.\footnote{Throughout this work
we use SNANA v10\_35g.}  Likelihoods are defined by comparing the
observed fluxes to model predictions in all passbands where the model
is defined.  In practice, this means we exclude the SN detections in
the F435W and F606W bands, which are too blue for our models at
$z>0.85$.

@@

The models shown in Figure~\ref{fig:lensingtest} are separated along
the $y$ direction into two broad classes, with
eight \change{so-called} {\it parametric}\footnote{\change{Although
this nomenclature is becoming standard in the literature, it is
somewhat misleading, as the pixels or grid cells in free-form models
are effectively parameters as well. Perhaps ``simply-parameterized''
would be more accurate, though we adopt the more common usage here.}}
models on top and six {\it free-form} models on the bottom.  Broadly
speaking, the parametric models use parameterized density
distributions to describe the arrangement of mass within the cluster,
typically associating massive dark matter halos with the positions of
bright cluster member galaxies. Therefore, parametric models rely (to
varying degrees) on the assumption that the 
\changeR{cluster's dark matter can be described
by analytic forms such as NFW halos \citep{Navarro:1997}, or pseudo isothermal elliptical mass distributions \citep{Kassiola:1993}.}
Free-form models divide the cluster into a grid, generally using a
\change{multi-scale grid} to get better sampling in denser regions. Each grid cell
is assigned a mass or a potential, and then the mass values are
iteratively refined to match the observed lensing
constraints. \change{In some cases an adaptive grid is used so
that the grid spacing itself can also be modified as the model is
iterated \citep[e.g.][]{Liesenborgs:2006,Merten:2009}.}

\change{
In general, free-form methods tend to explore the model parameter
space more fully than parametric methods.  This is reflected in the
size of the error-bars in Figure~\ref{fig:lensingtest}.  For example,
the Williams model (using GRALE), which has the largest, and possibly the
most realistic uncertainties is able to account for mass sheet
degeneracy by introducing a free parameter in the reconstruction that
represents an arbitrarily scaled mass sheet, with normalization to
be chosen by the genetic algorithm.  More complex degeneracies are
also accounted for in this model through the use of basis
functions \citep{Liesenborgs:2006,Liesenborgs:2007,Mohammed:2014} }

@@

This shift to higher magnifications is not, however, unique to the
SN \tomas\ position.  \citet{Jauzac:2014c} noted that their updated
mass model of Abell 2744 results in a systematic increase in the
magnification values across the cluster field. Examining a sample of
$\sim$30 multiple images, \citeauthor{Jauzac:2014c} found that the
magnifications increased by a factor of $\sim1.5$ relative to pre-HFF
models.  Although this single sight-line can only provide a very
limited test of the model, the \changeG{$4.3\sigma$} discrepancy is at
least a strong suggestion that \changeR{even with a large number of
high-quality} strong-lensing constraints \changeR{it is still possible
to generate} magnification maps that are susceptible to systematic
biases.

@@

The other three models in our comparison set also use the strong
lensing constraints from the core region, but additionally utilize
weak lensing measurements to provide additional constraints on the
mass distribution farther from the cluster core.  The signal from weak
lensing relies on a large sample of background galaxies, so this
constraint operates principally at separations more than 1\arcmin\
from the cluster core.  At a projected separation of $\sim40\arcsec$,
SN \tomas\ falls in between the strong- and weak-lensing regimes, and
one would expect that models incorporating both of those constraints
would be less susceptible to the \changeR{possible} systematic bias of
a sharply steepening mass profile.  The lowest magnification estimate
in our sample (from the Merten model) is among this sub-sample.
However, we find that collectively these three strong+weak models
(shown as diamonds in Figure~\ref{fig:lensingtest}) exhibit the same
propensity to overestimate the magnification along this line of sight.
Thus, a sudden change in the mass profile outside the strong lensing
region is not a likely explanation for this small systematic bias.

@@

This single object behind a single cluster is not in and of itself a
cause for alarm.  Previous analyses of lensed \SNIa\ found no
significant discrepancy beween the observed \SNIa\ magnifications and
the predictions from lens models
(\citetalias{Patel:2014}; \citealt{Nordin:2014}), albeit with a much
smaller set of lens models being tested.  The observed systematic bias
for \tomas\ is small, and many of the lens models being evaluated are
preliminary models that have not been updated to include all of the
HFF data.  Future revisions of the lens models for Abell
2744 \changeR{could} either incorporate the observed magnification
of \tomas\ as a new model constraint, or can revisit this test to
evaluate whether the bias persists.
